Assignment 2: Interpreting Quantitative Findings

### 2022-12-12 paper_to_html.Rmd
final edits

---
title: 'Assignment 2: Interpreting Quantitative Findings'
subtitle: |
  | 
  | University of Glasgow
  | `r source("scripts/utils.R")`
  | Student ID: 2819052
  | Course: Quantitative Methods
  | Number of words: 2680 
  | `r # maybe for word count: RmdWords("paper.Rmd")$num_words`
output:
  pdf_document:
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
  html_document: default
header-includes:
- \usepackage{amsmath} # format math stuff
- \usepackage{booktabs} # makes table nicer
- \usepackage{lastpage} # be able to count total pages
- \usepackage{fancyhdr} # header/footer
- \pagestyle{fancy} # header/footer
- \fancyhead{} # remove header
- \renewcommand{\headrulewidth}{0pt} # remove header
- \fancyfoot[CO,CE]{Page \thepage\ of \pageref{LastPage}} # add Page X of Y
- \usepackage{floatrow} # position tables and figures correctly
- \floatsetup[figure]{capposition=top} # position tables and figures correctly
- \floatsetup[table]{capposition=top} # position tables and figures correctly
bibliography: references.bib # add bibliography
csl: chicago-author-date # choose chicago citation style
link-citations: yes # add links in citations
nocite: '*' # add all used citations
---

@@

In 1998, the Good Friday/Belfast Agreement was signed and later validated by the voters, and this agreement marked a new era of hope towards peace, equality, and inclusion in Northern Ireland [@galligan2013gender]. The main focus on the agreement and referendum was obviously to stop the violent conflict and seek peace, but one part of the agreement also emphasized an equality agenda [@Hayward2021]. Thus, the agreement also marked the first Northern Irish formal recognition of women's rights to political inclusion [@galligan2013gender]. However, a formal recognition does not necessarily imply that gender equality trickles down into social norms and practices. Therefore, this report examines the contemporary state of women's equality in Northern Ireland.

@@

The causal relationship that we examine is thus a negative relationship. Although, we call this relationship causal, it is important to clarify that I do not imply that there is anything biological or deterministic that reduce women's income in general. Rather, the causal link is interpreted as a result of the societal discriminatory practices explained in the previous paragraph. To make a convincing inquiry of this causal relationship, we must also control for a number of other factor related to income and gender. This is important to make sure that our analysis do not simply show a spurious correlation as it could in fact be another factor that influences both gender and income. These control variables are introduced in [Data and Method].

<!--  \pagebreak -->
# Data and Method
<!-- Word count: approx. 800 -->
<!-- Word count: approx. 1158 -->

@@

Our sample for the analysis consists of data from Northern Ireland Life and Times Survey (NILT). The data have been collected every year since 1998, but this analysis uses the survey from 2012 [@ark_2012]. The respondents for the NILT survey was chosen from a systematic random sample of addresses. From this sample of 2126 addresses, 1204 questionnaires were fully completed using partly face-to-face- and self-completion questionnaires [@ark_2012]. Our sample of analysis is further reduced as not all respondents have answered our variables of interest, and we are therefore not able to include those respondents in our analysis. In the tables below, we present descriptive statistics for our variables of interest in both the full NILT sample and our sample of analysis. The first tables shows the categorical variables, where a frequency distribution are shown [@fogarty2018quantitative 88]. The second table shows the numerical variables, where we have calculated the mean and standard deviation [@fogarty2018quantitative 95]. For a full visualization of all variables included in the analysis, see [Appendix].

<!-- TC:ignore -->
```{r echo = FALSE, results='asis'}
sampling_table_categorical()
sampling_table_numerical()
```
<!-- TC:endignore -->

@@

The dependent variable or outcome that we want to explain in this analysis is income. Although income in itself is not complex concept compared to other theoretical concepts in social science, it can still be measured in many different ways. In this analysis, we measure income by the question: *What is your personal income before tax and national insurance contributions? (£ per year)*. It is thus a numeric, most specifically a ratio scale with an value of absolute zero - although it is quite uncommon to have a personal income of zero [@fogarty2018quantitative 56].

The independent variable or explanatory variable of our analysis is gender. Gender is a much more complex concept, and there is a large discussion about the distinction between gender and sex, and the biological and cultural components of these [@unger1993sex]. Quantitative research have mostly been dominated by gender being measured as a binary division between male and female. Thus, we are also limited to such a measurement of gender in the NILT 2012 survey. In this analysis, we therefore measure gender by a nominal, binary variable of either "male" or "female".

@@

There are several insights from this visualization. The vertical differences between the descriptive statistics, mean and median signifies that there is a general income difference between males and females, and that females in average have lower incomes than men. Initially, this suggests an acceptance of our alternative research hypothesis, but we will wait will any conclusion till the analysis, where we also control for other related factors. Another interesting insight is that there the distribution of income seems to slightly skewed. For both males and females, there are some respondents with a very high income compared to most other respondents. This is also seen as there is a difference between the mean and the median for both groups. These respondents have such extreme high incomes that they might be outliers compared to the general trend, and we need to be aware of that in our analysis  [@fogarty2018quantitative 221-222].

@@

<!--  \pagebreak -->
# Results and discussion
<!-- Word count: approx. 1300 -->
<!-- 735 -->

This report examines the question of whether gender affect income in contemporary Northern Ireland. As described in the [Introduction], the analysis therefore seeks to accept whether the null hypothesis or the alternative hypothesis - that being a woman rather than a man affects a lower income. As described in the [Data and Method] section, our analysis consists of survey data, where our main variables are answered gender and personal income. To examine our hypothesis, the following table shows the results of two regressions. First, our model is shown as a bivariate regression excluding our control variables, and secondly our full model is shown including our control variables.

<!-- TC:ignore -->
```{r, echo=FALSE, results='asis'}
source("scripts/regression.R")
```
<!-- TC:endignore -->

The primary finding from the regression results is that the variable gender have an significant effect on income. We see that the p-value is less than 0.01. This indicates that there is indeed a significant difference between the income of men and women. However, a significant p-value does not by itself means that the difference is important or meaningful - we also need to interpret the effect size [@field2012discovering 57]. If we take a look at the effect size ($\beta_1$), it has a value of approx -5,068. Thus according to the model, women in average earns 5,068 pounds a year less than men. That seems quite substantial as the average income in our sample is 16,892 pounds a year (See Table 2). Based on our analysis, we therefore accept our alternative hypothesis - that being a women rather than a man has a significant negative effect in your income.

In model, where we include our control variables, we see that there is trade union membership and being a supervisor has a significant positive effect on income, whereas age and to some extent also having a homosexual orientation rather than a heterosexual orientation have a significant negative effect on income. However, the effect of gender is independent of religion, sexual orientation, constitutional view, trade union membership, being a supervisor, and age, as including these control variables into the regression does not render the effect of gender on income insignificant. If we compare the model with and without control variables, we see that both R^2 and adjusted R^2 is increased with indicates that including the control variables does indeed yield a better fit of data.

@@

Another concern to our model is heteroscedasticity, i.e. whether there is not a constant variance of residuals on the levels of the dependent variable, income [@field2012discovering 272]. In the plot below, we visualize the fitted values with the residuals to see if the variances is constant along the x-axis. Below the plot is also shown the results of a Breusch-Pagan test to more formally test to heteroscedasticity [@fogarty2018quantitative 228].

@@

Unfortunately, the plot shows that the variance seem to increase for higher values of our dependent variables, income, and the Breusch-Pagan test also suggests that there is heteroscedasticity as the p-value is lower than 0.001. One solution to this problem could be to re-estimate our regression model with robust standard errors, but that it has not been done for this report [@fogarty2018quantitative 229].

<!-- \pagebreak -->
# Conclusion
<!-- Word count: approx. 500 -->
<!-- 361 -->

>>>>

---
title: 'Assignment 2: Interpreting Quantitative Findings'
subtitle: |
  | 
  | University of Glasgow
  | `r source("scripts/utils.R")`
  | Student ID: 2819052
  | Course: Quantitative Methods in the Social Sciences
  | Number of words: 2747
  | `r # maybe for word count: RmdWords("paper.Rmd")$num_words`
output:
  html_document: 
    number_sections: yes
    toc: true
    toc_depth: 2
  pdf_document:
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
header-includes:
- \usepackage{amsmath} # format math stuff
- \usepackage{booktabs} # makes table nicer
- \usepackage{lastpage} # be able to count total pages
- \usepackage{fancyhdr} # header/footer
- \pagestyle{fancy} # header/footer
- \fancyhead{} # remove header
- \renewcommand{\headrulewidth}{0pt} # remove header
- \fancyfoot[CO,CE]{Page \thepage\ of \pageref{LastPage}} # add Page X of Y
- \usepackage{floatrow} # position tables and figures correctly
- \floatsetup[figure]{capposition=top} # position tables and figures correctly
- \floatsetup[table]{capposition=top} # position tables and figures correctly
- \usepackage{setspace} # controls line spacing
bibliography: references.bib # add bibliography
csl: chicago-author-date # choose chicago citation style
link-citations: yes # add links in citations
nocite: '*' # add all used citations
---

@@

\doublespacing <!-- OBS: We use double spacing here. Could also be changed to \onehalfspacing -->

In 1998, the Good Friday/Belfast Agreement was signed and later validated by the voters, and this agreement marked a new era of hope towards peace, equality, and inclusion in Northern Ireland [@galligan2013gender]. The main focus on the agreement and referendum was obviously to stop the violent conflict and seek peace, but a part of the agreement also emphasized an equality agenda [@Hayward2021]. Thus, the agreement also marked the first Northern Irish formal recognition of women's rights to political inclusion [@galligan2013gender]. However, a formal recognition does not necessarily imply that gender equality trickles down into societal norms and practices. Therefore, this report examines the contemporary state of women's equality in Northern Ireland.

@@

The causal relationship that we examine is thus a negative relationship. Although, we call this relationship causal, it is important to clarify that I do not imply that there is anything biological or deterministic that reduce women's income in general. Rather, the causal link is interpreted as a result of the societal discriminatory practices explained in the previous paragraph. To make a convincing inquiry of this causal relationship, we must also control for a number of other factors related to income and gender. This is important to make sure that our analysis do not simply show a spurious correlation as it could in fact be another factor that influences both gender and income. These control variables are introduced in [Data and Method].

\pagebreak
# Data and Method
<!-- Word count: approx. 800 -->
<!-- Word count: approx. 1158 -->

@@

Our sample of analysis consists of data from the Northern Ireland Life and Times Survey (NILT). The data have been collected every year since 1998, but this analysis uses the survey from 2012 [@ark_2012]. The respondents for the NILT survey was chosen from a systematic random sample of addresses. From this sample of 2126 addresses, 1204 questionnaires were fully completed using partly face-to-face- and self-completion questionnaires [@ark_2012]. Our sample of analysis is further reduced as not all respondents have answered our variables of interest, and we are therefore not able to include those respondents in our analysis. In the tables below, we present descriptive statistics for our variables of interest in both the full NILT sample and our sample of analysis. The first tables shows the categorical variables, where a frequency distribution are shown [@fogarty2018quantitative 88]. The second table shows the numerical variables, where we have calculated the mean and standard deviation [@fogarty2018quantitative 95]. For a full visualization of all variables included in the analysis, see [Appendix].

<!-- TC:ignore -->
```{r echo = FALSE, results='asis'}
sampling_table_categorical("html")
sampling_table_numerical("html")
```
<!-- TC:endignore -->

@@

The dependent variable or outcome that we want to explain in this analysis is income. Although income in itself is more concrete concept compared to other theoretical concepts in social science (such as e.g. "solidarity"), it can still be measured in many different ways. In this analysis, we measure income by the question: *What is your personal income before tax and national insurance contributions? (£ per year)*. It is thus a numeric, more specifically a ratio scale with a value of absolute zero - although it is quite uncommon to have a personal income of zero [@fogarty2018quantitative 56].

The independent variable or explanatory variable of our analysis is gender. Gender is a much more complex concept, and there is a large discussion about the distinction between gender and sex, and the biological and cultural components of these [@unger1993sex]. Quantitative research have mostly been dominated by gender being measured as a binary division between male and female. Likewise, we are also limited to such a measurement of gender in the NILT 2012 survey. In this analysis, we therefore measure gender by a nominal, binary variable of either "male" or "female".

@@

There are several insights from this visualization. The vertical differences between the descriptive statistics, mean and median signifies that there is a general income difference between males and females, and that females in average have lower incomes than men. Initially, this suggests an acceptance of our alternative research hypothesis, but we will wait with any conclusion till the analysis, where we also control for other related factors. Another interesting insight is that the distribution of income seems to be slightly skewed. For both males and females, there are some respondents with a very high income compared to most other respondents. This can also noticed as there is a difference between the mean and the median for both groups. These respondents have such extreme high incomes that they might be outliers compared to the general trend, and we need to be aware of that in our analysis  [@fogarty2018quantitative 221-222].

@@

\pagebreak 
# Results and discussion
<!-- Word count: approx. 1300 -->
<!-- 735 -->

This report examines the question of whether gender affect income in contemporary Northern Ireland. As described in the [Introduction], the analysis therefore seeks to accept whether the null hypothesis or the alternative hypothesis - that being a woman rather than a man affects a lower income. As described in the [Data and Method] section, our analysis consists of survey data, where our main variables are gender and personal income. To examine our hypothesis, the following table shows the results of two regressions. First, our model is shown as a bivariate regression excluding our control variables, and secondly our full model is shown including our control variables.

<!-- TC:ignore -->
```{r, echo=FALSE, results='asis'}
source("scripts/regression_to_html.R")
```
<!-- TC:endignore -->

The primary finding from the regression results is that the variable gender have an significant effect on income. We see that the p-value is less than 0.01. This indicates that there is indeed a significant difference between the income of men and women. However, a significant p-value does not by itself means that the difference is important or meaningful - we also need to interpret the effect size [@field2012discovering 57]. If we take a look at the effect size ($\beta_1$), it has a value of approx -5,068. Thus according to the model, women in average earns 5,068 pounds a year less than men. That seems quite substantial as the average income in our sample is 16,892 pounds a year (See Table 2). Based on our analysis, we therefore accept our alternative hypothesis - that being a woman rather than a man has a significant negative effect in your income.

In the model, where we include our control variables, we see that trade union membership and being a supervisor has a significant positive effect on income, whereas age and to some extent also having a homosexual orientation rather than a heterosexual orientation have a significant negative effect on income. However, the effect of gender is independent of religion, sexual orientation, constitutional view, trade union membership, being a supervisor, and age, as including these control variables into the regression does not render the effect of gender on income insignificant. If we compare the model with and without control variables, we see that both $R^2$ and adjusted $R^2$ is increased with indicates that including the control variables does indeed yield a better fit of data.

@@

Another concern to our model is heteroscedasticity, i.e. whether there is not a constant variance of residuals on the levels of the dependent variable, income [@field2012discovering 272]. In the plot below, we visualize the fitted values with the residuals to see if the variances is constant along the x-axis. Below the plot is also shown the results of a Breusch-Pagan test to more formally test for heteroscedasticity [@fogarty2018quantitative 228].

@@

Unfortunately, the plot shows that the variance seem to increase for higher values of our dependent variables, income, and the Breusch-Pagan test also suggests that there is heteroscedasticity as the p-value is lower than 0.001. One solution to this problem could be to re-estimate our regression model with robust standard errors, but this has not been possible within the scope of this report [@fogarty2018quantitative 229].

\pagebreak
# Conclusion
<!-- Word count: approx. 500 -->
<!-- 361 -->
